Death By Dropdown?

### 2025-04-03 conference-materials/2025_shiny-conf/slides.qmd
Refines speaker notes for clarity and engagement
- Streamlines speaker notes to enhance presentation flow and audience interaction.
- Focuses on simplifying language and emphasizing key psychological principles
and their application to dashboard design.
- Clarifies the BID framework's
benefits and practical implementation using the {bidux} package.

Updates enhance the practical impact of the presentation and improve audience engagement with concise, relatable examples.

- Start by acknowledging the shared pain point everyone recognizes
- Emphasize the emotional response users have when facing cluttered dashboards
- Mention that this is costing organizations real money through delayed decisions
- "Dashboard rot" = dashboards that get built but never used
- Ask for a show of hands: "How many of you have abandoned using a dashboard because it was too complex?" (engage audience early)
:::

@@

- The central thesis of our talk: storytelling vs. data dumping
- Introduce SoundPulse as our case study - they're a composite client...competitor of Spotify
- Their market research team needed to present competitor analysis to executives
- Emphasize that this shift is both attitudinal and technical
- Preview: "We'll show you exactly how this transformation works using real code examples"
- When the GIF appears, pause briefly - people will laugh at the "drowning in paper" visual
:::

@@

::: {.notes}
**Show Impact**: Build up intuition of annoyance/frustration using "before" dashboard.

- Even thought SoundPulse is a composite client...this UI is actually a fair representation of what many of our market research clients have come to expect from their dashboards
  * We didn't just pick the worst dashboard we could find...we again, put a composite together to give you a representative idea of the kinds of challenges we've faced as devs trying to help our clients with the products we deliver
- Walk through the key issues with the "before" dashboard:
  * "Look at all those filters - users had to make decisions before seeing ANY data"
  * 8 tabs of dense data tables...need to know what you want to figure out in advance
  * "No guidance on where to start - cognitive overload from the beginning"
  * No one wants to use a dashboard like this...and no one will...UIs like this are a major contributor to dashboard rot
:::

@@

- The 5 stages form an acronym: NISAV (pronounced "knee-sav")
- Each stage builds on psychological principles but translates to practical actions
- Emphasize this isn't just theory - it's a practical workflow for developers
- Preview: "We'll walk through each stage with SoundPulse's case study"
- Mention these are iterative, not strictly linear - you can revisit earlier stages
- Personal anecdote: "We developed this framework after seeing the same patterns of dashboard failure across industries"
:::

@@

::: {.notes}
**Context Setting**: Position the framework within the broader field.

- Acknowledge that this builds on existing work - "We're standing on the shoulders of giants"
- Emphasize the unique integration of fields - psychology, UX, data & visual communication
  * Quickly mention some influences: "Edward Tufte's data visualization principles, Kahneman's work on cognitive biases, etc."
- Key difference from other approaches: process-oriented vs. checklist approach
- "What makes BID different is that it's specifically designed for analytic applications (like dashboards)"
  * The ultimate goal of the BID framework is to helping end users gain clearer insights from dashboards...while giving developers a structured approach to building more effective applications
:::

@@

::: {.notes}
**Case Study Introduction:**

- "So...let's take a look at how we used this framework to help SoundPulse refactor their dashboard."
- **Stage 1 (Notice):** The first thing we noticed...as I'm sure you all of you did as well...was that users were overwhelmed with options - dropdown menus and complex tables
  * This is what we call 'cognitive load' - when the brain gets overtaxed with information.
  * We also identified a violation of Hick's Law, which simply means that more choices lead to longer decision times and confusion.
- **Stage 2 (Interpret):** Once we identified these issues, we focused on something called 'Processing Fluency' - making information easier to digest.
  * This meant using color-coding and visual patterns that the brain can process instantly without conscious effort.
  * You'll see in our code examples in a moment just how we implemented this with color scales in reactable.
- **Stage 3 (Structure):** "We then restructured using the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements. 
  * Now there are a ton of ways you can do this
  * As you'll see in our code example, we redesigned the layout to visually group metrics that belong together, and highlight the metrics that mattered most to create a natural flow through the data
- Now you don't need to know all of these concepts to use the framework...we'll come back to that
  * Don't need to be a psychologist to use the framework
- But first Milena is going to show you *how* we translated these concepts into code...
:::

@@

::: {.notes}
**Visual Impact**: Connect UI to stages/concepts just mentioned.

- "Here's what our reactable implementation looks like in practice"
- For each image as it appears:
  * First image: "Notice how the grouped bars implement Stage 1 principles by reducing cognitive load"
  * Second image: "The dynamic legend addresses Hick's Law by simplifying choices"
  * Third image: "Color coding enhances processing fluency from Stage 2"
  * Fourth image: "The grouping clearly implements the Principle of Proximity from Stage 3"
- "Each of these visual elements directly ties to a principle from the BID framework"
- "This is the power of intentional design - every element has a psychological purpose"
:::

@@

::: {.notes}
**Basic Implementation**: Start with the foundational elements.

- "Let's break down the simplest implementation of BID principles in reactable"
- **WOW MOMENT #1**: "The groupBy parameter does double duty here - it addresses cognitive load from Stage 1 while creating visual hierarchy for Stage 3"
- "Color coding isn't just for aesthetics - it directly implements processing fluency from Stage 2"
- "Notice how the conditional styling creates instant visual cues for good and concerning metrics"
- "This simple implementation already reduces cognitive load and enhances processing - but we can do more"
- Transition: "Now let's add more advanced features to handle Stage 3 principles..."
:::

@@

::: {.notes}
**Advanced Features**: Highlight sophisticated techniques.

- "Now we're implementing more advanced BID principles with reactable"
- **WOW MOMENT #2**: "The columnGroups feature implements Dual-Processing Theory by separating quick insights (System 1) from detailed analysis (System 2)"
- "The create_bar_chart_cell function creates custom HTML for enhanced bar charts with tooltips"
- **WOW MOMENT #3**: "Progressive disclosure with the details parameter is one of the most powerful cognitive load reduction techniques - users only see the details when they need them"
- "This nested reactable implementation lets users drill down into detailed data without being overwhelmed"
- "All of these features work together to create a dashboard that respects how humans actually process information"
- Ask audience: "How many of you have used columnGroups or details in reactable?" (interactive moment)
- Transition: "Now let's see how these principles look in the final implementation..."
:::

@@

::: {.notes}
**Later stages:** Setup for echarts examples

- "For the later stages of the BID framework, we addressed more subtle psychological factors that affect how users interpret and act on data."
- **Stage 4 (Anticipate):** "We noticed executives were making flawed comparisons without proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * As you'll see in our echarts4r example, we added dynamic benchmark lines to provide meaningful context
  * We also implemented a 'Framing Toggle' to address how different teams focus on different perspectives on the same data."
- **Stage 5 (Validate):** "Finally, we applied the 'Peak-End Rule' - the principle that people judge experiences mainly by their most intense point and how they end.
  * Our code example shows how we created summary cards that highlight key takeaways, ensuring users walk away with the most important insights."
- Let's take a closer look at that code now...
:::

@@

::: {.notes}
**Complete Dashboard**: Show the integrated result.

- "Here's the complete dashboard implementing multiple BID principles with echarts4r"
- For each image as it appears:
  * First image: "The gauge charts provide instant comprehension while reference lines provide anchoring"
  * Second image: "The framing toggle allows different perspectives on the same data"
  * Third image: "The summary cards implement the Peak-End Rule from Stage 5"
- "Notice how the consistent visual design implements the Beautiful-Is-Good Stereotype from Stage 5"
- "The card-based layout creates clear visual hierarchy and reduces cognitive load"
- "Every element in this dashboard has a specific psychological purpose"
- "This is what we mean by behavior-informed design - intentional choices based on how humans think"
- Transition: "Now let's look at how framing affects decision-making in more detail..."
:::

@@

::: {.notes}
**Gauge Charts**: Start with visually impactful elements.

- "Let's start with gauge charts that implement processing fluency principles from Stage 2"
- **WOW MOMENT #5**: "The e_gauge function creates visually intuitive displays that are instantly comprehensible"
- "Notice how we're customizing the gauge to eliminate unnecessary elements like the pointer"
- "The color gradient visually reinforces the value - showing progress as filled area"
- "Custom tooltips provide additional context to mitigate confirmation bias from Stage 4"
- "This function encapsulates a reusable pattern that we can apply consistently across metrics"
- "The visual simplicity of gauges reduces cognitive load while maintaining information richness"
- Transition: "Now let's add reference benchmarks to address the Anchoring Effect..."
:::

@@

- "This code directly implements the Anchoring Effect principle from Stage 4"
- **WOW MOMENT #6**: "The e_mark_line feature provides visual reference points that anchor interpretation"
- "Without these reference lines, values like 65% or 45% have no context - are they good or bad?"
- "By providing benchmarks, we're giving users appropriate anchors for comparison"
- "Notice the styling choices - orange dashed line creates visual distinction from the data"
- "The tooltip provides additional context about the benchmark source"
- "This simple addition dramatically improves how users interpret the data"
- Transition: "Now let's look at our most sophisticated implementation - the framing toggle..."
:::

@@

::: {.notes}
**Framing Toggle**: Demonstrate psychological impact of framing.

- "This is our most powerful implementation of a BID principle - the Framing Toggle"
- **WOW MOMENT #7**: "We're creating two different views of the same data to demonstrate how framing affects perception"
- "The 'Progress' frame emphasizes achievement - how far we've come"
- "The 'Gap' frame emphasizes what remains - how far we still need to go"
- "This directly implements the Framing & Loss Aversion principle from Stage 4"
- "The JavaScript allows users to switch between frames without a full page reload"
- "Different teams at SoundPulse used different frames - marketing used Progress, product used Gap"
- "This is a powerful example of how design choices influence decision-making"
- Interactive moment: "Let me show you how this works in practice..." (if possible, demonstrate live)
- Transition: "Now let's see how these elements come together in the complete dashboard..."
:::

@@

::: {.notes}
**Framing Effect**: Emphasize psychological impact.

- "This slide demonstrates the power of framing in action - same exact data, completely different psychological impact"
- "Progress framing creates motivation and highlights achievement - 'We're 65% there!'"
- "Gap framing creates urgency and focuses on what remains to be done - '35% still to go!'"
- "This psychological principle has been demonstrated in countless studies"
- "Present a 90% survival rate vs. a 10% mortality rate - people react completely differently"
- "At SoundPulse, this had real organizational impact"
    - Marketing team preferred progress framing for stakeholder presentations, while product teams found gap framing more actionable for development priorities.
- "Marketing needed to show progress to build confidence with stakeholders"
- "Product teams needed to focus on gaps to prioritize development resources"
- "The toggle feature meant both teams could use the same dashboard for different purposes"
- "This is why understanding psychological principles is so powerful for dashboard design"
- Interactive moment: "Think about your own dashboards - are you framing data in a way that drives the right decisions?"
- Transition: "Now let's talk about how we're making these principles accessible to every developer..."
:::

@@

::: {.notes}
**Package Introduction**: Generate excitement about the tooling.

- "We're not just sharing a framework - we're building tools to make it easy to adopt"
- "We're excited to introduce `{bidux}`...our new R package currently in **Phase 1 development** that helps implement the BID framework with your workflows"
- "Our goal is to make these psychological principles accessible to every R developer"
- Current features focus on documentation and knowledge
- Future features will include ready-to-use UI components with BID principles built in
- "The LLM integration will let you describe what you want in plain language"
- "Think of {bidux} as your behavioral science assistant for Shiny app development"
:::

@@

::: {.notes}
**Code Example**: {bidux} introduction

- Let's look at our Phase 1 implementation of the {bidux} package, which we're releasing today.
- The concept dictionary provides case-insensitive lookups with implementation tips - not just theory, but practical advice on applying each concept.
- You can search across psychology, UI/UX, and data visualization concepts to find what's most relevant to your problem.
- The core of the package is a pipeline of functions that guide you through the 5 BID stages:
  * `bid_notice()` now auto-suggests appropriate theories based on your problem description
  * `bid_interpret()` helps structure your data story with enhanced details
  * `bid_structure()` documents your layout decisions and related concepts
  * Each function provides helpful suggestions as you work
- When we add the upcoming component suggestion system, you'll get tailored UI recommendations based on your BID analysis - showing exactly how to implement these principles in your Shiny code.
- This is just Phase 1 - in Phases 2-4, we'll be adding LLM integration and ready-to-use UI components.
- When the link appears: "We're launching this at ShinyConf today - check out the repo"
:::

@@

::: {.notes}
**Results Summary**: Emphasize concrete benefits.

- These outcomes reflect consistent feedback across client organizations that have embraced our framework. The qualitative data tells a compelling story:
  * (PAUSE) Teams are making decisions with much greater speed and confidence
  * (PAUSE) Users are discovering more actionable insights in each session
  * (PAUSE) And perhaps most telling—stakeholders have shifted from being skeptical to becoming advocates
- We're currently building telemetry into our dashboards to gather more precise measurements...but the message from our users is already clear: this framework is changing how teams interact with their data
- When the testimonial appears: "This kind of transformation is what BID is all about"
- "Remember, this wasn't just visual changes - this was a fundamental shift in how our users...and developers...approached the dashboard"

- "While we didn't collect formal metrics before and after implementation, the qualitative feedback was tremendous. SoundPulse reported dramatically faster decision-making, more insights per session, and a complete transformation in stakeholder satisfaction - from frustration to enthusiasm."
:::

@@

- "BID is about reducing friction between users and insights"
- "The tools we've shown today - reactable and echarts4r - give you everything you need"
- "{bidux} is available now to help you get started - check out our GitHub"
- "Most importantly, remember this isn't just about pretty dashboards..."
- "It's about transforming how you and others make decisions with data"
- "Every dropdown you eliminate, every chart you clarify, helps someone make a better decision"
- "This is about making your users successful, not just making dashboards"
- Transition: "Thank you for your attention - we're excited to take your questions"
:::

@@

- Thank the audience sincerely
- Remind them of key resources: "Remember to check out the GitHub repo at github.com/bidux"
- "We'll be around after the session if you want to chat more"
- "We'd love to hear about your dashboard challenges and how BID might help"
- For common questions, prepare brief answers:
  * "How do you measure success?" → "Start with user testing and session analytics"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too"
  * "How long does implementation take?" → "Start small - even one principle makes a difference"
- "Don't forget to follow us on LinkedIn/GitHub for updates on {bidux} development"
:::

>>>>

- "Raise your hand if you've ever encountered a dashboard that looked like this..." (gesture to image)
- "Dashboard rot" = dashboards built but never used - costing real development time and money
- Ask: "How many of you have abandoned using a dashboard because it was too complex?" (engage audience)
:::

@@

- Key message: "The answer isn't more data - it's better storytelling with data"
- "SoundPulse, a music streaming analytics team, faced this exact problem"
  * They're a composite client
- Their market research team needed to present competitor analysis to executives
  * Typically involves turning tons of complex analyses into non-technical insights
- "We're going to show you exactly how psychological principles transformed their dashboard - with approaches and code you can use immediately"
:::

@@

::: {.notes}
**Show Impact**: Build frustration recognition with the "before" dashboard.

- Even though SoundPulse is a composite client...this is representative of real dashboards we've seen across the market research industry
- Point out key issues:
  * "Filter overload before seeing ANY insights"
  * "Multiple tabs with no clear starting point"
  * "Complete cognitive overload"
- "This is why dashboards get abandoned - nobody wants this experience"
:::

@@

- "The 5 stages form an easy workflow: Notice → Interpret → Structure → Anticipate → Validate"
  * Click through each
- "Each stage translates psychological principles into practical developer actions"
- "This isn't just theory - it's a step-by-step process you can implement today"
- "These stages are iterative - you'll often revisit earlier stages as you discover new insights"
:::

@@

::: {.notes}
**Context Setting**: Position the framework within the broader fields.

- "BID integrates existing best practices from multiple disciplines"..."We're truly standing on the shoulders of giants here"
- Key differentiator: "But what makes BID unique is its process-oriented framework vs. a typical checklist approach"
- "The goal is helping users gain clearer insights while giving developers a tool to use in their workflows"
- "Even though there's a lot of theory, you don't need a psychology degree to apply these principles effectively"
- "To demonstrate...let's take a closer look at how we connected these stages to SoundPulse's dashboard."
:::

@@

::: {.notes}
**Case Study Introduction:** Keep energy high while introducing concepts.

- **Stage 1 (Notice):** The first thing was users were overwhelmed with options
  * This is what we call 'cognitive load' - when the brain gets overtaxed with information
  * We was a violation of Hick's Law, which simply means that more choices lead to longer decision times and confusion
- **Stage 2 (Interpret):** We then focused on something called 'Processing Fluency' - making the information easier to digest
  * For example, using color to create a visual hierarchy of information so the brain can process it more easily
- **Stage 3 (Structure):** "We then restructured the UI using the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements together. 
  * Now there are a ton of ways you can do this...
  * But as you'll see in our code example, we created groups of primary and secondary metrics to create a natural flow through the data
- Now we'll show you the actual code implementation of these principles
:::

@@

::: {.notes}
**Visual Impact**: Connect visuals to principles just mentioned.

- For each image as it appears:
  * First image: "Notice how grouping reduces cognitive load"
  * Second image: "The consistent legend simplifies comprehension"
  * Third image: "Color coding creates instant visual processing"
  * Fourth image: "Proximity principles group related metrics"
- "This is the power of intentional design - every element has a psychological purpose"
:::

@@

::: {.notes}
**Basic Implementation**: Highlight the simplicity of applying BID.

- "This first example shows how just a few lines implement powerful BID principles"
- "groupBy reduces cognitive load while creating visual hierarchy"
- "Color coding isn't just aesthetic - it creates instant processing fluency"
- "Progressive disclosure means users only see details when needed"
- "These three techniques alone dramatically improve dashboard usability"
- "This simple implementation already reduces cognitive load and enhances processing - but we can do more"
:::

@@

::: {.notes}
**Advanced Features**: Highlight sophisticated techniques of package, but balance with BID principles.

- "columnGroups implements Principle of Proximity - grouping primary and secondary metrics together"
- "The dynamic legend reduces choice complexity - addressing Hick's Law"
- "Color coding by demographic creates instant pattern recognition"
- "Notice how each feature directly maps to a BID principle"
- Ask: "How many have used columnGroups in reactable before?" (engagement opportunity)
  * If not many...great, we taught something!
  * If a lot...could also mention (quickly) how to extend with legend/div
:::

@@

::: {.notes}
**Later stages:** Set up the psychological depth of later stages

- "For the later stages of the BID framework, we addressed more subtle psychological factors that affect how users interpret and act on data."
- **Stage 4 (Anticipate):** We noticed executives were making flawed comparisons without proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, so we addressed this with something called a "framing toggle"...more on that in just a moment
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule' - the idea that people judge experiences mainly by their most intense point and how they end
  * We leveraged this by creating summary cards that highlight key takeaways, increasing the chances that users leave the app with a clear, memorable impression of the most important insights
:::

@@

::: {.notes}
**Complete Dashboard**: Show integration of multiple principles

- For each image as it appears:
  * First image: "Gauge charts provide instant comprehension"
  * Second image: "The framing toggle changes perspective while maintaining data integrity"
  * Third image: ""Summary cards implement the Peak-End Rule"
- "Also, the card-based layout creates clear visual hierarchy and reduces cognitive load"
- "Every element in this dashboard has a specific psychological purpose"
- "This is what we mean by behavior-informed design - intentional choices based on how humans think"
:::

@@

::: {.notes}
**Gauge Charts**: Focus on visual processing benefits.

- "Gauge charts implement processing fluency perfectly"
- "We're customizing the gauge to eliminate unnecessary complexity"
- "The color gradient visually reinforces the value"
- "This visual simplicity reduces cognitive load while maintaining information richness"
- "The function encapsulates this pattern for consistent application"
:::

@@

- This directly addresses the Anchoring Effect from Stage 4"
- "Without reference lines, metrics like '65%' lack meaning - is it good or bad?"
- "Benchmarks provide the context users need for proper interpretation...and helps with alignment across users"
- "The distinctive orange dashed line creates visual separation from the data"
- "This simple addition dramatically improves decision quality"
:::

@@

::: {.notes}
**Framing Toggle**: Emphasize the powerful psychological effect.

- "This implementation shows how framing affects perception - same data, different decisions"
- "The 'Progress' frame emphasizes achievement - how far we've come"
- "The 'Gap' frame emphasizes what remains - how far we still need to go"
- "Different teams at SoundPulse used different frames - marketing used Progress, product used Gap"
- "The server-side code dynamically recalculates values based on the framing perspective"
- "This is psychological design directly impacting business decisions"
:::

@@

::: {.notes}
**Framing Effect**: Drive home the practical impact.

- This is a great example to illustrate the theory behind these concepts
- "In this case we are showing the *Same exact data*...but users experience completely different psychological impact"
- "Progress framing creates motivation and highlights achievement"
  - "Marketing preferred progress framing to build stakeholder confidence"
- "Gap framing creates urgency and focuses on what remains to be done"
  - "Product teams used gap framing to prioritize development"
- "This isn't just theory - a classic JDM study presented a 90% survival rate vs. a 10% mortality rate - people react completely differently"
- "This is why understanding psychological principles is so powerful for dashboard design"
- "Think about your own dashboards - are you framing data in a way that drives the right decisions?"
:::

@@

::: {.notes}
**Package Introduction**: Generate excitement about practical tools.

- "We're not just sharing a framework or theories - we're giving you tools to implement it immediately"
- "The `{bidux}` package makes these principles accessible to every R developer"
- "Phase 1 is available today on GitHub"
- Future phases will include LLM integration and ready-to-use UI components"
- "Think of `{bidux}` as your behavioral science assistant for Shiny app development"
:::

@@

::: {.notes}
**Code Example**: Show practical usage of the package

- "Let's look at what {bidux} offers today"
- "The concept dictionary provides implementation advice, not just theory"
- "You can search across psychology, UI/UX, and data visualization fields...and more!"
- "The workflow functions guide you through each BID stage"
- "Auto-suggestion of theories based on your problem description"
- "Component suggestions provide concrete UI implementation ideas"
- When the link appears: "We're launching this at ShinyConf today - check out the repo for full documentation"
:::

@@

::: {.notes}
**Results Summary**: Emphasize real-world impact.

- "The results speak for themselves"
- "Users who previously abandoned the dashboard now rely on it daily"
- "Decisions that took days now happen in minutes"
- We're currently building telemetry into our dashboards to gather more precise measurements...but the message from our users is already clear: this framework is changing how teams interact with their data
- When testimonial appears: "This transformation is what BID delivers"
- "Remember, this wasn't just about visual improvements - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::

@@

- "BID reduces friction between users and insights"
- "The tools we've shown today give you everything you need to get started"
- "`{bidux}` is available now to help structure your implementation"
- "Every dropdown you eliminate helps someone make a better decision"
- "This is about making your users successful, not just making dashboards"
:::

@@

- We'd love to hear about your dashboard challenges"
- "Don't forget to follow us on LinkedIn/GitHub for updates on {bidux} development"
- "Feel free to connect with us for any questions or deeper discussions!"
- Quick answers to likely questions:
  * "How do you measure success?" → "Start with user testing and session analytics"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too"
  * "How long does implementation take?" → "Start small - even one principle makes a difference"
:::
